---
title: "Problem Set 8"
author: "Adam Ten Hoeve"
date: '`r format(Sys.time(), "%d %B, %Y")`'
header-includes:
  - \usepackage{amsmath}
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

These questions were rendered in R markdown through RStudio (<https://urldefense.com/v3/__https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf__;!!NCZxaNi9jForCP_SxBKJCA!H2o4AVX3C2X-TmYBjS04rM_qVhCWeHQ0mcn9vyoT7AvlzUgSuJFJ379LaZsNInZefg$ >, <https://urldefense.com/v3/__http://rmarkdown.rstudio.com__;!!NCZxaNi9jForCP_SxBKJCA!H2o4AVX3C2X-TmYBjS04rM_qVhCWeHQ0mcn9vyoT7AvlzUgSuJFJ379LaZtOrdH-oA$ > ).

Please complete the following tasks, in R where applicable. Please generate a solution document in R markdown and upload the rendered .doc, .docx, or .pdf document.  You may add hand computations to a .doc or .docx if you prefer. In the rendered document, please show your code. That is, don't use "echo=FALSE".  

In either case, your work should be based on the data's being in the same folder as the R files. Please turn in your work on Canvas. Your solution document should have your answers to the questions and should display the requested plots.

## 1. (5 points) 

Let $X$ and $Y$ be independent random variables with $\chi^2$ distributions with $n$ and $m$ degrees of freedom respectively. What is the distribution of $X+Y$?. (5 points)

We know that for $X$ and $Y$ to be $\chi^2$ random variables, they would result from the sum of $n$ and $m$, respectively, iid standard normal distributions. In other words, if $Z_{i,j} \sim N(0,1)$ for $i\in \{1, 2\}$ then $X = Z_{1, 1} + \dots + Z_{1,n}$ and $Y = Z_{2,1} + \dots + Z_{2,m}$.

Now we can see that our sum $X+Y$ is actually the sum of many iid standard normal distributions, i.e. $X + Y = Z_{1,1} + \dots + Z_{1,n} + Z_{2,1} + \dots + Z_{2,m}$, which would be distributed by another $\chi^2$ distribution with $n+m$ degrees of freedom.
 
## 2.

Correlation may be reported as a measure of the linear association between two variables. An understanding of typical correlations between independent variables aids in understanding these reports.

Consider two iid samples $X$ and $Y$, each of size $n$, from a standard Normal distribution. The following question investigates the range of correlations that are typical for such samples.

### 2a.(10 points)

Perform a numerical experiment to estimate the  $(0.025,0.25,.0.5, 0.75,975)$ quantiles of the sample correlations between two iid samples of sizes 10, 20, 50, and 100 from the standard Normal distribution. Display the quantiles in a table. Suggestion: for each size, simulate the process of drawing a pair of samples calculating their correlation many times then report the requested quantiles of the observed correlations.(5 points)

```{r}
set.seed(0)

sample_correlation = function(sample_size){
    # Generate random samples from the standard normal
    sample1 = rnorm(sample_size)
    sample2 = rnorm(sample_size)
    # Return the correlation between the samples
    return(cor(sample1, sample2))
}

n = 5000
sample_sizes = c(10, 20, 50, 100)
quantiles = c(0.025, 0.25, 0.5, 0.75, 0.975)
df = data.frame(matrix(0, nrow=length(sample_sizes), ncol=length(quantiles)))

index = 1
# For each sample size, simulate the correlation
for (sample_size in sample_sizes){
    # Generate n correlations from std.norms of the sample_size
    cors = replicate(n, sample_correlation(sample_size))
    
    # Calculate the quantiles for the correlations
    sample_quantiles = quantile(cors, probs=quantiles)
    # Add the quantiles to the dataframe
    df[index, ] = sample_quantiles
    index = index + 1
}

names(df) = quantiles
row.names(df) = sample_sizes
df
```

### 2b.(10 points)

Display the quantiles for each $n$ in a plot that has the value of $n$ on the x-axis, and the correlation quantiles on the y-axis. Please use bars to indicate the range between the 0.25 and 0.75 quantile and the range between the 0.025 and 0.975 quantile. Please plot a point at the 0.5 qunatile for each $n$. Please title the plot and label the axes with informative text. You may find "geom_errorbar" useful. You can position ticks on the x-axis using "scale_x_continuous(breaks=c(10,20,50,100))" for example.

```{r}
g = ggplot(df, aes(x=c(1, 2, 3, 4)))

g + 
    # Add errorsbars for [0.025, 0.975] range
    geom_errorbar(aes(ymin=df$"0.025", ymax=df$"0.975", colour="blue"), width=0.2) +
    # Add errorbars for [0.25, 0.75] range
    geom_errorbar(aes(ymin=df$"0.25", ymax=df$"0.75", colour="red"), width=0.2) + 
    # Add points for the 0.5 quantile
    geom_point(aes(x=c(1, 2, 3, 4), y=df$"0.5")) +
    # Change the xaxis label and ticks
    scale_x_discrete(limits=c("10", "20", "50", "100")) +
    # Change the title and axis labels
    labs(title="Quantiles of Correlations among Std. Normal Samples of Different Sample Sizes", 
         x="Sample Size", y="Correlation Quantile")
```



## 3. (5 points each)

The data in "pew_sampled_data.RData" come from the PEW RESEARCH CENTER
2017 STEM SURVEY downloaded from https://urldefense.com/v3/__https://www.pewsocialtrends.org/dataset/2017-pew-research-center-stem-survey/__;!!NCZxaNi9jForCP_SxBKJCA!H2o4AVX3C2X-TmYBjS04rM_qVhCWeHQ0mcn9vyoT7AvlzUgSuJFJ379LaZupcRJ_5w$ 
8/5/2019.

The data here are sampled from the full survey data according to the survey weights.

```{r}
load("pew_sampled_data-2.RData")
```

### 3a.

The variable recording whether the respondent was a STEM worker or a non-STEM worker is "WORKTYPE_FINAL".

The variable "SCH7" records the respondents' selection based on the following question and options:

What's the main reason many young people don't pursue college degrees in science, technology, engineering and mathematics? Is it mostly because. 

1	They think these subjects are too hard

2	They think these subjects are too boring

3	They think these subjects are not useful for their careers

4	Some other reason (please specify) 

5 Refused

Please create a 2-way table that gives the proportion of respondents among Stem workers that selected each response and the proportion of respondents among non-Stem workers that selected each response. 

For the purposes of table display, please shorten the responses to "too hard", "too boring", "not useful", "other".

```{r}
# Calculate the count of each grouping of worktype and response
counts = samp %>% 
            group_by(WORKTYPE_FINAL, SCH7) %>% 
            summarize(n = n())

# Format the counts into a 2-way table
prop.table = matrix(c(counts$n[1:3], sum(counts$n[4:5]), counts$n[6:8], sum(counts$n[9:10])), ncol=2)
two_way = data.frame(t(prop.table))
names(two_way) = names=c("too hard", "too boring", "not useful", "other")
row.names(two_way) = c("stem", "nonstem")
two_way

round(two_way/rowSums(two_way), 4)
```
 
  
## 3b.

Please do a $\chi^2$  test and a Fisher's Exact Test to determine if the responses to "SCH7" are independent of "WORKTYPE_FINAL", and compare the results. 
  
```{r}
# Chi-squared test on two_way data
chisq.test(two_way)
# Fisher test on two_way data
fisher.test(two_way)
```
Both tests have very small p-values, each $\sim 0.02$ meaning that we should reject the null hypothesis and assert that there is some statistically significant difference between the STEM and NON-STEM groups.


## 3c.

Please repeat 2b. omitting respondents who answered "refused".

```{r}
chisq.test(samp$WORKTYPE_FINAL[samp$SCH7 != "Refused"], 
            samp$SCH7[samp$SCH7 != "Refused"])

fisher.test(samp$WORKTYPE_FINAL[samp$SCH7 != "Refused"], 
            samp$SCH7[samp$SCH7 != "Refused"])
```
When we remove the "Refused" answers, we get even smaller p-values than when we left them in, which leads us to the same conclusion as in part b. We should reject the null hypothesis and assert that there is some statistically significant difference between the STEM and NON-STEM groups.

### Harassment

The variable "HARASS3" records the respondents' selection based on the following question and options:

Have you ever personally experienced sexual harassment at work, or have you not experienced this?

1	Yes, I have experienced sexual harassment at work

2	No, I have not experienced sexual harassment at work

3 Refused

(Note that the responses are self-reported. Respondents may differ in their perception of harassment and their willingness to report harassment.)

The genders of the respondents are recorded as "Male" or "Female" in the variable "PPGENDER"

(Note that limitation to binary gender reports makes these data ineffective for study of the association of non-binary gender identification with the remaining variables. )

One categorization of the ethnicities of the respondents is recorded in the variable "PPETHM"

### 3d.

Please create a subset of the data that omits responses from the small categories "HARASS3"=="Refused". The function "droplevels" applied to a factor variable will omit levels that aren't represented in the data.

Please run and report the result of an appropriate test of the null hypothesis that the response to "HARASS3" is independent from the response to "PPGENDER" on this subset. Please give a statistical interpretation of the results.

```{r}
# Remove any data tuples where HARASS = "Refused"
samp.harass = samp[samp$HARASS3 != "Refused", ]

# To determine independence, we can use a chi-squared test
chisq.test(samp.harass$HARASS3, samp.harass$PPGENDER)

# Let's also check with the fisher test
fisher.test(samp.harass$HARASS3, samp.harass$PPGENDER)
```
Both of our tests resulted in a very small p-value, therefore we should reject the null hypothesis, and should assert that the HARASS3 and PPGENDER variables are not independent.

### 3e.

Please restrict the data from 3a the respondents who report "Female" in the "PPGENDER" variable. Please provide a table of the proportion of each category of "PPETHM" that responds "Yes, I have experienced sexual harassment at work". Please run and report the result of an appropriate test of the null hypothesis that the response to "HARASS3" is independent from the response to "PPETHM" on this subset. Please give a statistical interpretation of the results. 

```{r}
# Restrict the data to just females
samp.female = samp.harass[samp.harass$PPGENDER == "Female", ]

# Construct a 2-way test for the proportions
counts.female = samp.female %>%
                 group_by(PPETHM, HARASS3) %>%
                 summarize(n = n())

two.way.female = data.frame(matrix(head(counts.female$n, -1), ncol=2, byrow=TRUE),
               row.names=c("White", "Black", "Other", "Hispanic"))
names(two.way.female) = c("Yes", "No")
# Print out the values and proportions
two.way.female
round(two.way.female / rowSums(two.way.female), 4)

# Conduct some tests to see if HARASS3 and PPETHM are independent
chisq.test(two.way.female)
fisher.test(two.way.female)

```

Both tests result in a very small p-value, therefor we should reject the null and assert that the HARASS3 and PPETHM variables are not independent.
